On the effects of partying on next-day sports performance

### 2017-05-31 writing/draft.tex
Formatting and polishing the first rough draft. Updated results table-names. Added a count of # of games in dataset.

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.\cite{hoch}
Within the NBA, J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
A few interviews with former professional athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife. We exploit data on bookmaker spreads, the expected score 
differential between two teams after conditioning on observable performance. 
We expect a team to meet the spread half the time, since this is how bookmakers 
minimize risk on their end. We construct a model which attempts to 
estimate the causal effect of  visiting a ``party-city'' on subsequent day 
performance as measured by the spread. Since next-day
opponent is uncorrelated with exposure to treatment, we have identification in our variable of 
interest.

@@

\section{Materials and Methods}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to determine when players partied, one might consider to look at social media data
for direct mentions of partying, either by athletes or witnesses. However, due to
intense public scrutiny that athletes face, there is a strong reluctance to post
such mentions on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s trip to Miami in 2017 season.\cite{bleler}
The reports are few and far enough between,
and whether they are even reported at all is strongly biased on whether a particular city 
has nightclubs with VIP rooms, etc; picking up statistical evidence would be fruitless in this case.

\paragraph{Two approaches: a discrete indicator and a continuous measure of nightlife}
Because famous athletes often enjoy a celebrity lifestyle, using conventional statistics
such as number of drinking establishments in each city does not work; athletes instead
tend to visit high end nightclubs. We again turned to player interviews for ideas,
and found two major insights. From our conversations, 
we obtained a listing of cities which were anecdotally
mentioned as having active night-life. Between players, the rank ordering of these
was consistent to the extent that LA and NY were top
offenders. This motivates using a binary indicator for party in these two cities. 
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a
simple count of the number of musicians in the county in which the stadium resides. 
We pursue both approaches.

@@

For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically over-valued
teams which you are more likely to play after a party city,
we could see a spurious effect. 
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reliable. So, if we do find an effect in our variable of interest,
we can be confident in our conclusion that there is a causal effect and not some
other systematic bias.

@@

\paragraph{Fatigue from Jet lag}
What about jet-lag? Prolonged air travel has in general been shown
to have a negative effect on sports performance.\cite{leeandgalvez}
Previous studies have shown that jet lag is a strong enough effect
to erase home field advantage in MLB.\cite{songetal}

@@

\subsection{Data}
\paragraph{Lines data} We first look toward 
\href{http://www.basketball-reference.com/leagues/}{basketball-reference.com}
for a listing of game-days by season. With this listing, we queried \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{covers.com},
for all games occurring on a particular game, and for each we observe
outcome of each game alongside the line set by a betting house. We obtain
data for seasons 2010-11 through 2016-17.

@@

\paragraph{Location and Travel Distance Data} We are also interested
to account for the effects of jet-lag. For each team, we queried
a maps search engine to obtain a longitude and latitude listing.\cite{ggmap}
For each pairwise set of coordinates, we calculated distance between the two
points.\cite{sp} To account for jet-lag, we first computed the final direction 
(bearing) when arriving at the 
second point after starting from the first and following the shortest path on a sphere 
(following a great circle), which yields a direction in degrees. 
We then obtain a measure for east-west travel by looking at the trigonometric-$\sin$ of this angle.
Interacting this with travel-distance gets us a measure of jetlag.\cite{lallensack}

@@

\subsection{Robustness check}
As a robustness check, we verify that this phenomena persists in an entirely separate
sport with entirely different players. To that end, we look toward
Major League Baseball, which is also well set up for our quasi-experiment since
there are many games played back-to-back in the season (more in fact, than NBA).

In general, we can replicate our analysis, but there are also some notable differences.
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. The logic
is that because there are so many more games played in MLB, the weekend effect is more salient.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

% Table created by stargazer v.5.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu % Date and time: Mon, May 29, 2017 - 11:56:17 PM
\hspace{-105pt}
\begin{tabular}{@{\extracolsep{5pt}}lcccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{4}{c}{\textit{Dependent variable:}} \\  \cline{2-5}  \\[-1.8ex] & \multicolumn{2}{c}{NBA: Meet Spread} & Demeaned Reb/Min & MLB: Win Game \\  \\[-1.8ex] & \multicolumn{2}{c}{\textit{logistic}} & \textit{OLS} & \textit{logistic} \\  \\[-1.8ex] & (1) & (2) & (3) & (4)\\  \hline \\[-1.8ex]  party.discrete & $-$0.406$^{***}$ (0.133) &  &  &  \\    party &  & $-$0.053$^{**}$ (0.022) & $-$0.001$^{*}$ (0.0005) & $-$0.035$^{*}$ (0.020) \\    lag.chg.pos & 0.0001 (0.002) & 0.0001 (0.002) & 0.00003 (0.0001) &  \\    odds &  &  &  & 2.602$^{***}$ (0.162) \\    log(travel.dist + 1) & $-$0.017 (0.025) & $-$0.015 (0.025) & 0.001 (0.001) & $-$0.001 (0.005) \\    ew & $-$0.114 (0.230) & $-$0.101 (0.230) &  &  \\    nhours.lgame & 0.001 (0.001) & 0.001 (0.001) & 0.00003 (0.00002) &  \\    I(hour(date)) & 0.007 (0.012) & 0.005 (0.012) &  &  \\    weekend &  &  &  & 0.040 (0.036) \\    I(team == location) & $-$0.011 (0.043) & $-$0.001 (0.043) &  & 0.025 (0.034) \\    log(travel.dist + 1):ew & 0.018 (0.035) & 0.016 (0.035) &  &  \\    ndays.lgame &  &  &  & 0.014 (0.019) \\    Constant & $-$0.077 (0.374) & $-$0.041 (0.374) & $-$0.007$^{*}$ (0.004) & $-$1.331$^{***}$ (0.080) \\   \hline \\[-1.8ex]  Observations & 10,000 & 9,979 & 44,190 & 25,215 \\  R$^{2}$ &  &  & 0.0003 &  \\  Adjusted R$^{2}$ &  &  & 0.0002 &  \\  Log Likelihood & $-$6,925.180 & $-$6,912.606 &  & $-$17,274.450 \\  Akaike Inf. Crit. & 13,868.360 & 13,843.210 &  & 34,562.900 \\  Residual Std. Error &  &  & 0.105 (df = 44185) &  \\  F Statistic &  &  & 3.064$^{**}$ (df = 4; 44185) &  \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

>>>>

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.\cite{hoch}
Within the NBA, J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
Matt Barnes recently complained that ``there's no nightlife in Utah'' when 
asked about playing Jazz in the second round of NBA playoffs in 2017.\cite{cestone}
A few interviews with former professional NBA athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

We are interested to see if 
teams exhibit a decline in performance the day following a game in a city 
with active nightlife. We exploit data on bookmaker spreads, the expected score 
differential between two teams after conditioning on observable performance. 
We expect a team to meet the spread half the time, since this is how bookmakers 
minimize risk on their end. We construct a model which attempts to 
estimate the causal effect of  visiting a ``party-city'' on subsequent day 
performance as measured by the spread. To the extent that next-day
opponent is uncorrelated with exposure to treatment, we have identification in our variable of interest.

@@

\section{Materials and Methods}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to determine when players partied, one might consider to look at social media data
for direct mentions of partying, either by athletes or witnesses. However, due to
intense public scrutiny that athletes face, there is a strong reluctance to disclose
their personal lives on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s recent trip to Miami.\cite{bleler}
However, these reports are few and far enough between,
and whether they are even reported at all is strongly biased on whether a particular city 
has nightclubs with VIP rooms, etc; picking up statistical evidence would be fruitless in this case.

\paragraph{Two approaches: a discrete indicator and a continuous measure of nightlife}
Because famous athletes often enjoy a celebrity lifestyle, using conventional statistics
such as number of drinking establishments in each city does not work; athletes instead
tend to visit high end nightclubs. We again turned to player interviews for ideas,
and found two major insights. From our conversations, 
we obtained a listing of cities which were anecdotally
mentioned as having active night-life. Between players, the rank ordering of these
was consistent to the extent that Los Angeles and New York city were top
offenders. This motivates using a binary indicator for party in these two cities. 
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a
simple count of the number of musicians in the county in which the stadium resides. 
We pursue both approaches.

@@

For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically over-values
teams which are more likely to be encountered after playing a city with active nightlife,
we could see a spurious effect. 
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reasonably reliable. 
So, if we do find an effect in our variable of interest,
we can be confident in our conclusion that there is a causal effect and not some 
other systematic bias.

@@

\paragraph{Fatigue from jetlag}
What about jetlag? Prolonged air travel has in general been shown
to have a negative effect on sports performance.\cite{leeandgalvez}
Previous studies have shown that jetlag is a strong enough effect
to erase home field advantage in major league baseball, for example.\cite{songetal}

@@

\subsection{Data}
\paragraph{Lines data} We first look toward 
\href{http://www.basketball-reference.com/leagues/}{basketball-reference.com}
for a listing of game-days by season. With this listing, we queried \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{covers.com},
for all games occurring on a particular day, and for each we observe the
outcome of each game alongside the line set by a betting house.\cite{hrisports} We obtain
data for seasons 2010-11 through 2016-17. We are missing the first couple months of games 
in the 2010-11 season, and our 2011-12 season has some 229 missing games. In each remaining
season, we have at most two games missing of the $82 \times 30 / 2 = 1,230$ played.
We have a total of 7,829 games in our data-set.

@@

\paragraph{Location and Travel Distance Data} We are also interested
to account for the effects of jet-lag. For each team, we queried
a maps search engine to obtain a longitude and latitude listing.\cite{ggmap}
For each pairwise set of coordinates, we calculated distance between the two
points.\cite{sp} To account for jet-lag, we first computed the final direction 
(bearing) when arriving at the 
second point after starting from the first and following the shortest path on a sphere 
(following a great circle), which yields a direction in degrees. 
We then obtain a measure for east-west travel by looking at the trigonometric-$\sin$ of this angle after converting to radians.
Interacting this with travel-distance gets us a measure of jetlag.\cite{lallensack}

@@

\subsection{Robustness check}
As a robustness check, we verify that this phenomena persists in an entirely separate
sport with entirely different players. To that end, we look toward
Major League Baseball, which is also well set up for our quasi-experiment since
there are many games played back-to-back in the season (more in fact, than NBA).
There are also anecdotes of players celebrating with alcohol in public.\cite{miller,campbell}

In general, we can replicate our analysis and use the same resources to obtain our data,
 but there are also some notable differences.
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. The logic
is that because there are so many more games played in MLB, the weekend effect is more salient.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

% Table created by stargazer v.5.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu % Date and time: Mon, May 29, 2017 - 11:56:17 PM
\hspace{-105pt}
\begin{tabular}{@{\extracolsep{5pt}}lcccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{4}{c}{\textit{Dependent variable:}} \\  \cline{2-5}  \\[-1.8ex] & \multicolumn{2}{c}{NBA: Meet Spread} & Demeaned Reb/Min & MLB: Win Game \\  \\[-1.8ex] & \multicolumn{2}{c}{\textit{logistic}} & \textit{OLS} & \textit{logistic} \\  \\[-1.8ex] & (1) & (2) & (3) & (4)\\  \hline \\[-1.8ex]  party.discrete & $-$0.406$^{***}$ (0.133) &  &  &  \\    party.continuous &  & $-$0.053$^{**}$ (0.022) & $-$0.001$^{*}$ (0.0005) & $-$0.035$^{*}$ (0.020) \\    lag.change.position & 0.0001 (0.002) & 0.0001 (0.002) & 0.00003 (0.0001) &  \\    odds &  &  &  & 2.602$^{***}$ (0.162) \\    log(travel.dist+1) & $-$0.017 (0.025) & $-$0.015 (0.025) & 0.001 (0.001) & $-$0.001 (0.005) \\    east-west.travel.dir& $-$0.114 (0.230) & $-$0.101 (0.230) &  &  \\    nhours.last.game & 0.001 (0.001) & 0.001 (0.001) & 0.00003 (0.00002) &  \\    hour(date) & 0.007 (0.012) & 0.005 (0.012) &  &  \\    weekend &  &  &  & 0.040 (0.036) \\    (team == location) & $-$0.011 (0.043) & $-$0.001 (0.043) &  & 0.025 (0.034) \\    log(travel.dist+1):ew & 0.018 (0.035) & 0.016 (0.035) &  &  \\    ndays.last.game &  &  &  & 0.014 (0.019) \\    Constant & $-$0.077 (0.374) & $-$0.041 (0.374) & $-$0.007$^{*}$ (0.004) & $-$1.331$^{***}$ (0.080) \\   \hline \\[-1.8ex]  Observations & 10,000 & 9,979 & 44,190 & 25,215 \\  R$^{2}$ &  &  & 0.0003 &  \\  Adjusted R$^{2}$ &  &  & 0.0002 &  \\  Log Likelihood & $-$6,925.180 & $-$6,912.606 &  & $-$17,274.450 \\  Akaike Inf. Crit. & 13,868.360 & 13,843.210 &  & 34,562.900 \\  Residual Std. Error &  &  & 0.105 (df = 44185) &  \\  F Statistic &  &  & 3.064$^{**}$ (df = 4; 44185) &  \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}
